### 2016-07-06 src/5_1_Proposal.tex
Adding first proposal

\chapter{Online VECM proposal}
Financial time series are known for their non-stationary behaviour. However,
sometimes they exhibit some stationary linear combinations. When this happens,
it is said that those time series are cointegrated. The Vector Error Correction
Model (VECM) is an econometric model which characterises the joint dynamic
behaviour of a set of cointegrated variables in terms of forces pulling towards
equilibrium. In this study, we propose an Online VEC model (OVECM) which
optimises how model parameters are obtained using a sliding window of the most
recent data. Our proposal also takes advantage of the long-run relationship
between the time series in order to obtain improved execution times. Our
proposed method is tested using four foreign exchange rates with a frequency of
1-minute, all related to the USD currency base. OVECM is compared with VECM and
ARIMA models in terms of forecasting accuracy and execution times. We show that
OVECM outperforms ARIMA forecasting and enables execution time to
be reduced considerably while maintaining good accuracy levels compared with
VECM.

\vspace{0.5cm}

\section{The problem}
In finance, it is common to find variables with a long-run and/or a short-run
equilibrium relationship. This relationship is called cointegration and it
reflects the idea of that some set of variables cannot wander too far away from
each other.

Cointegration means that one or more combinations of these variables is
stationary even though individually they are not.

Some financial models, such as  vector error correction model (VECM), take advantage of
this property and describe the joint behaviour of several cointegrated
financial instruments.

VECM is a special case of the vector autoregressive model (VAR). VAR is a linear
model which express future values in terms of its own history and other
variables past values. If cointegration exists, VECM allows to introduce an
error correction term due to its cointegration estimation error.

VECM as well as VAR model parameters are obtained using ordinary least squares
method (OLS). Since OLS involves many calculations, the parameter estimation
method is computationally expensive when the number of past values and
observations considered increases. Moreover, OLS is an ill-posed problem which
admits an infinite number of solutions.

Ridge regression method (RR) tackles this ill-posed problem and is usually
formulated instead of OLS. RR includes a regularisation parameter that leads to
better generalisation capability. However, RR is still computationally
expensive dealing with large datasets. Recently, online learning algorithms
have been proposed to solve problems with large data sets because of their
simplicity and their ability of updating the model when new data is available.

We propose an online formulation of the VEC model based on the
AAR method considering only a sliding window of the most recent data. The
algorithm introduces matrix optimisations in order to reduce the number of
operations. Our method is later tested with financial data from the foreign
exchange market.

\section{Methodology} \label{sec:methodology}

Since VECM parameter estimation is computationally expensive, we propose an
online version of VECM (OVECM).  OVECM considers only a sliding window of the
most recent data. Moreover, since cointegration vectors represent long-run
relationships which vary little in time, OVECM determines firstly if they require calculation.

OVECM also implements matrix optimisations in order to reduce execution time,
such as updating matrices with new data, removing old data and introducing new
cointegration vectors.

Algorithm~\ref{alg:proposal} shows our OVECM proposal which considers the
following:

\begin{itemize}
\item The function \texttt{getJohansen} returns cointegration vectors given by
the Johansen method considering the trace statistic test at 95\%
significance level.
\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
and~(\ref{eq:vecY}) that allows VECM to be solved.
\item The function \texttt{vecMatrixOnline} returns the
matrices~(\ref{eq:vecA}) and~(\ref{eq:vecY}) aggregating new data and removing
the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
\item Out-of-sample outputs are saved in the variables 
$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
\item The model is solved using OLS.
\item In-sample outputs are saved in the variables $\Delta
\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
series.
\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
MAPEs are stored in the variable $e$. If the average of MAPEs
($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
\item If new cointegration vectors are required, the function
\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
\end{itemize}

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: number of past values \\
$L$: sliding window size ($L<N$) \\
$\text{mean\_error}$: MAPE threshold \\
$n$: interpolation points to obtain MAPE \\
\ENSURE  $\,$ \\
$\{\mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
	\IF {i = 0 }
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
        \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
	\ELSE
	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{Y})$}
        \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
    \ENDIF
    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
    \STATE $e \gets \texttt{mape}(\mathbf{Y}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
    \IF {$\texttt{mean}(e) > \text{mean\_error}$}
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$\mathbf{A} \gets
        \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
    \ENDIF
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} \gets \mathbf{Y}[L+1:N]$
\STATE $\mathbf{Y}_{\text{pred}}\gets 
\mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
\end{algorithmic}
\caption{OVECM: Online VECM}
\label{alg:proposal}
\end{algorithm}

A pseudocode of the algorithm \ref{alg:proposal} is detailed in the appendix \ref{fig:pseudocodigo}.

Our proposal was compared against VECM and ARIMA. Both algorithms were adapted
to an online context in order to get a reasonable comparison with our proposal
(see algorithms \ref{alg:SLVECM} and \ref{alg:SLARIMA}). VECM and ARIMA are
called with a sliding window of the most recent data, whereby the models are
updated at every time step.

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: number of past values \\
$L$: sliding window size ($L<N$) \\
\ENSURE  $\,$ \\
$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1]$
        \STATE $model = VECM(\mathbf{y}_i, p)$
        \STATE $\mathbf{Y}_{\text{pred}}[i] = model.predict(\mathbf{y}[i+L])$
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N] $
\end{algorithmic}
\caption{SLVECM: Sliding window VECM}
\label{alg:SLVECM}
\end{algorithm}

Since we know out time series are I(1) SLARIMA is called with $d=1$. ARIMA is
executed for every time series.

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: autoregressive order \\
$d$: integrated order\\
$q$: moving average order\\
$L$: sliding window size ($L<N$) \\
\ENSURE  $\,$ \\
$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
\FOR { $j =0$ to $l-1$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1,j]$
        \STATE $model = ARIMA(\mathbf{y}_i, (p,d,q))$
        \STATE $\mathbf{Y}_{\text{pred}}[i,j] = model.predict(\mathbf{y}[i+L,j])$
\ENDFOR
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N,:] $
\end{algorithmic}
\caption{SLARIMA: Sliding window ARIMA}
\label{alg:SLARIMA}
\end{algorithm}

Both OVECM and SLVECM time complexity is dominated by Johansen method which is
$O(n^3)$. Thus, both algorithms order is $O(Cn^3)$ where $C$ is the number of
iterations.

\section{Results} \label{sec:results}

\subsection{Data} \label{sec:unitroot}
Tests of SLVECM, SLARIMA and our proposal OVECM were carried out using foreign
four exchange data rates: EURUSD, GBPUSD, USDCHF and USDJPY. This data was
collected from the free database Dukascopy which gives access to the Swiss
Foreign Exchange Marketplace ~\cite{Dukascopy2014}.

The reciprocal of the last two rates (CHFUSD, JPYUSD) were used in order to
obtain the same base currency for all rates.  The tests were done using
1-minute frequency from ask prices which corresponded to 1.440 data points per
day from the 11th to the 15th of August 2014.

\subsection{Unit root tests} \label{sec:unitroot}
Before running the tests, we firstly checked if they were I(1) time series
using the Augmented Dickey Fuller (ADF) test.

\begin{table}[h!]
\caption{Unit roots tests}
\label{tab:adf}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
\hline
EURUSD          & -0.64 & -1.94 & True       \\
$\Delta$EURUSD & -70.45   & -1.94 & False       \\
GBPUSD          & -0.63   & -1.94 & True          \\
$\Delta$GBPUSD & -54.53   & -1.94 & False       \\
CHFUSD          & -0.88   & -1.94 & True         \\
$\Delta$CHFUSD & -98.98   & -1.94 & False       \\
JPYUSD          & -0.65 & -1.94 & True        \\
$\Delta$JPYUSD & -85.78 & -1.94 & False     \\ 
\hline
\end{tabular}
\end{center}
\end{table}

Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
test but they rejected it with their first differences. This means that all of
them are I(1) time series and we are allowed to use VECM and therefore OVECM.

\subsection{Parameter selection} \label{sec:paramselection}

In order to set OVECM parameters: windows size $L$ and lag order $p$,
we propose to use several window sizes: $L = 100, 400, 700, 1000$. For every window size $L$ we chose lag order with minimum AIC.

ARIMA parameters were also obtained using AIC.
Parameters optimisation is presented in table~\ref{tab:params}:

\begin{table}[ht]
\caption{Parameters optimisation. VECM order and ARIMA parameters were selected
using AIC.}
\label{tab:params}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Windows size $L$ & VECM & ARIMA\\
 $L$ & order $(p)$ & order $(p,d,q)$ \\
\hline
100 & 2 & p=2,d=1,q=1\\
400 & 5 & p=1,d=1,q=1\\
700 & 3 &p=2,d=1,q=1\\
1000 &3 & p=2,d=1,q=1\\
\hline
\end{tabular}
\end{center}
\end{table}

In OVECM we also define a mean\_error variable, which was defined based on the
in-sample MAPEs. Figure \ref{fig:mapes} shows how MAPE moves and how mean\_error
variable help us to decide whether new cointegration vectors are needed.

\begin{figure}[!h]
  \centering
   \includegraphics[width = 7.5cm]{img/mapes} 
  \caption{In-sample MAPEs example for 50 minutes. The average of them is
  considered to obtain new cointegration vectors.}
  \label{fig:mapes}
 \end{figure}

\subsection{Execution times} \label{sec:exectimes}
We ran OVECM and SLVECM 400 iterations. SLARIMA execution time is excluded
because its is not comparable with OVECM and SLVECM. SLARIMA was created based
on statsmodels library routine ARIMA.

The execution times are shown in the table~\ref{tab:extimes}.

\begin{table}[h!]
\caption{Execution times}
\label{tab:extimes}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& $L$ & order & e  & Time[s] \\
\hline
OVECM & 100 &p=2  & 0      & 2.492\\
OVECM & 100 &p=2  & 0.0026  & 1.606\\
SLVECM & 100 &p=2& -- & 2.100\\
\hline
OVECM & 400 & p=5  & 0      & 3.513\\
OVECM & 400 &p=5  & 0.0041  & 2.569\\
SLVECM & 400 & p=5 & -- & 3.222\\
\hline
OVECM & 700 &p=3  & 0      & 3.296\\
OVECM & 700 &p=3  & 0.0032  & 2.856\\
SLVECM & 700 &p=3 & -- & 3.581\\
\hline
OVECM & 1000 & p=3 & 0      & 4.387\\
OVECM & 1000 & p=3  & 0.0022  & 2.408\\
SLVECM & 1000 & p=3  & -- & 3.609\\
\hline
\end{tabular}
\end{center}
\end{table}

It is clear that execution time depends directly on $L$ and $p$ since they
determine the size of matrix $\mathbf{A}$ and therefore affect the OLS function
execution time.
It is worthy of note that execution time also depends on mean\_error because it
determines how many times OVECM will recalculate cointegration vectors which is
an expensive routine.

Figure~\ref{fig:mapes} shows an example of the in-sample MAPE for 50 iterations.
When the average of the in-sample MAPEs is above mean\_error new cointegration
vectors are obtained. In consequence, OVECM performance increases when
mean\_error increases. However, this could affect accuracy, but
table~\ref{tab:stats} shows that using an appropriate mean\_error doesn't affect
accuracy considerable.

\subsection{Performance accuracy} \label{sec:performacc}

Table~\ref{tab:stats} shows in-sample and out-of-sample performance measures:
MAPE, MAE and RMSE for OVECM, SLVECM and SLARIMA. Test were done using the
parameters defined in table \ref{tab:params}.  We can see that OVECM has very
similar performance than SLVECM and this support the theory that cointegration
vectors vary little in time. Moreover, OVECM also out performed SLARIMA based on
these three performance measures.

\begin{table*}[ht!]
\caption{Model measures}
\label{tab:stats}
\begin{center}
\begin{tabular}{|l|l|c|c|c|c|c|c|c|c|}
\hline
\multicolumn{4}{|c|}{Model} & \multicolumn{3}{|c|}{In-sample} &
\multicolumn{3}{|c|}{Out-of-sample} \\ 
\hline
\hline
Method & $L$ & Parameters & e & 
MAPE & MAE& RMSE& 
MAPE & MAE& RMSE \\
\hline
 OVECM  &   100  &  P=2& 0.0026  &  0.00263&  0.00085&  0.00114&  0.00309&  0.00094&  0.00131\\
 OVECM  &   400  &  P=5& 0.0041  &  0.00378&  0.00095&  0.00127&  0.00419&  0.00103&  0.00143\\
 OVECM  &   700  &  P=3& 0.0032  &  0.00323&  0.00099&  0.00130&  0.00322&  0.00097&  0.00132\\
 OVECM  &   1000 &  P=3& 0.0022  &  
 \textbf{0.00175}&  \textbf{0.00062}&  \textbf{0.00087} &  
 \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
\hline
 SLVECM  &   100 &  P=2& -  &  0.00262&  0.00085&  0.00113&  0.00310&  0.00095&  0.00132\\
 SLVECM  &   400 &  P=5& -  &  0.00375&  0.00095&  0.00126&  0.00419&  0.00103&  0.00143\\
 SLVECM  &   700 &  P=3& -  &  0.00324&  0.00099&  0.00130&  0.00322&  0.00098&  0.00132\\
 SLVECM  &   1000 &  P=3& -  &  
 \textbf{0.00174}&  \textbf{0.00061}&  \textbf{0.00087}&  
 \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
\hline
SLARIMA & 100  &p=2, d=1, q=1 & - & 0.00285  &  0.00110  &  0.00308  &  0.00312  &0.00098  &  0.00144 \\
SLARIMA & 400  &p=1, d=1, q=1 & - & 0.00377  &  0.00101  &  0.00128  &  0.00418 & 0.00106 & 0.00145 \\
SLARIMA & 700  &p=2, d=1, q=1 & - & 0.00329  &  0.00102  &  0.00136  &  0.00324  & 0.00097  &  0.00133 \\
SLARIMA & 1000 &p=2, d=1, q=1 & - & \textbf{0.00281}  & \textbf{0.00074}  &  \textbf{0.00105}  &  \textbf{0.00177} & \textbf{0.00063}  &  \textbf{0.00092} \\
\hline
\end{tabular}
\end{center}
\end{table*}

>>>>

\chapter{Adaptive parallel algorithm for Vector error correction model}
VECM (presented in section \ref{sec:vecm}) has been
broadly used with low frequency time series. We aimed to adapt VECM to be used
in finance with high frequency stream data.

We observed that the number of cointegration relationships varies with the
length of historical data used. Moreover, parameters that increased these
relationships in time led to better forecasting performance. Our proposal,
called an Adaptive VECM (AVECM) is to make a parameters grid search that
maximises the number of cointegration relationships in the near past. To ensure
the search can be executed fast enough, we used a distributed environment.

The methodology was tested using four 10-second frequency time series of the
Foreign Exchange market. We compared our proposal with ARIMA and the naive
forecast of the random walk model. Numerical experiments showed that on average
AVECM performed better than ARIMA and random walk. Additionally, AVECM
significantly improved execution times with respect to its serial version.

\section{Introduction}
\label{sec:introduction}
In finance, it is common to find variables with long-run equilibrium
relationships. This is termed cointegration and reflects the idea that some sets
of variables cannot wander too far from each other. Cointegration means that one
or more linear combinations of these variables are stationary even though
individually they are not. Some models, such as the Vector Error Correction
(VECM), see \cite{engle87}, take advantage of this property and describe the
joint behaviour of several cointegrated variables. VECM introduces this long-run
relationship among a set of cointegrated time series as an error correction
term. These time series must be integrated of order 1, denoted I(1), i.e. they
become stationary at their first differences. In finance, I(1) time series are
very common and to introduce cointegration restrictions in models often improves
forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications, among others: \cite{mukherjee1995}, \cite{seong2013},
\cite{maysami2000} and \cite{arestis2001}. VECM has also been used in pair
trading, see \cite{herlemont2003}, or models with more than two variables, see
for example \cite{mukherjee1995} and \cite{engle2004}).

VECM parameters are obtained using the ordinary least squares (OLS) method,
developed by \cite{golub1980}. Since OLS involves many calculations, the
parameter estimation is computationally expensive when the number of lagged
values and data increases. Moreover, obtaining cointegration vectors is also an
expensive routine. This is a main limitation to use VECM with stream data with
high frequency. \cite{chen2003} addresses the advantage of distributed
processing over conventional rolling window processing.

Our aims were to study if a parallel version of VECM can be used with high
frequency stream data.  Our approach was to determine adaptively the number of
observations and lags of VECM which maximise cointegration relations in the
past.  This search is computationally expensive because the Johansen method, see
\cite{johansen1995}, is required to find cointegration vectors, which is of
order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error
Correction (AVECM), is to parallelise this search in order to get new parameters
before new data arrives.  Model effectiveness is focused on out-of-sample
forecast rather than in-sample fitting. This criterion allows AVECM prediction
capability to be expressed rather than just explaining data history.  The
forecast capability of our method was measured using MSE and the Theil's
$U$-statistic, see \cite{theil1966}, widely used in economic forecast. Tests
were run using four currency rates: Euro (EUR) to United States Dollar (USD)
(EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF)
and USD to Japanese Yen (JPY) (USDJPY) with a 10-second frequency.

\section{Methodology} \label{sec:methodology}
\subsection{Motivation} \label{sec:proposal}

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, VECM assumes cointegration vectors
do not change in time.  In fact, \cite{gregoryETal1996} addresses that the
long-run relationships between the time series  might change due to several
economic factors that can lead to structural breaks in the cointegration
relationship.  In order to show that the number of cointegration vectors depends
on the amount $L$ of historical data and the number of lags $p$ in the VECM, we
used a grid search.  We defined a grid of possible values for $L$ and $p$. $L$
goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a step size of
4 hours and $p$ throughout $[1,5]$ with step size of 1. The first experiment
consisted in determining the number of cointegration vectors for all
combinations of $L$ and $p$.  We used four forex rates: EURUSD, GBPUSD, USDCHF
and USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of
August 2014, when the New York and London financial markets opened.

Figure \ref{fig:hists} shows the distribution of the number of cointegration
vectors given by the Johansen method for different values of $L=[2,6,10,14]$
hours  and $p=1$ . This procedure was carried out by a sliding window of
historical data moving 1000 times.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.9\textwidth]{img/Fig1}
  \caption{Distribution of the number of cointegration vectors using $p=1$ lags.
  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
  hour = 360 data points).}
  \label{fig:hists}
\end{figure}

From section \ref{sec:background} we know that $r=0$ means no cointegration and
$r=l$ (we are using four rates, so $l=4$) reveals that no process is I(1) but
stationary.  The interesting cases of cointegration are those where $r$ lies
strictly between $0$ and $4$, i.e. $0<r<4$.

In order to measure the extent of cointegration, we introduce a
{\em percentage of cointegration\/} as following:
\begin{equation} \label{eq:pcoint}
PC = 
\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
     {\#it}\times 100
\end{equation}
where c.v. stands for cointegration vectors and $it$ is the number of iterations.

The goal of our next experiment was to find a relation between this ratio $PC$
and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
between $[700,1500]$ data points with step size 100 and $p$ took values between
$[1,5]$ with step size 1.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.9\textwidth]{img/Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Best windows size $L$ found was 1060. Below MSE versus $L$ shows a
  rapidly decreasing behaviour, founding minimum at PC = 80\% }
  \label{fig:cointvsmse}
\end{figure}

Figure~\ref{fig:cointvsmse} shows the relation between MSE and $PC$ and $L$.  We
found that better cointegration percentage leads to better performance accuracy
in terms of reducing MSE.

Therefore we proposed to choose $L$ and $p$ in order to maximise the percentage
of cointegration $PC$ in the near past. This process was done every time that
new data was processed. However, this search can be slow if we try different
values for $L$ and $p$ and therefore we distributed this calculation in order to
reduce searching time.

Our proposal is then a modified version of VECM, with parameter $p$ and the
amount of historical data used $L$ obtained at every step. Only the search of
$L$ and $p$ was done in a distributed environment, since this is the most
expensive routine. Our proposal is called Adaptive Vector Error Correction model
(AVECM). AVECM is detailed in the algorithm \ref{alg:AVECM} which summarises
our proposal.

\begin{algorithm}[ht!]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$j$: Starting point of testing \\
$it$: Ending point of testing \\
$ps$: list of $p$ values \\
$Ls$: list of $L$ values ($L<N$) \\
$m$: Iterations to determine parameters ($m < N-L$)\\
\ENSURE  $\,$ \\
$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
\FOR { $i =j$ to $it$ }
   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
    \STATE $L,p \gets
    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
\ENDFOR
\end{algorithmic}
\caption{AVECM: Adaptive VECM.}
\label{alg:AVECM}
\end{algorithm}

The input of AVECM is time series prices which are cointegrated. The starting
point of testing is $j$ and the total number of iterations is $it$. We need to
ensure that $j$ is at least the maximum value of $Ls$. $Ls$ and $ps$ are the
possible values for $L$ and $p$.  The function {\bf get\_best\_params} makes
this grid search on the two vector lists $Ls$ and $ps$ and returns the
parameters $L$ and $p$ which maximise the percentage of cointegration $PC$ (see
equation~\ref{eq:pcoint}) for a pre-defined number of iterations $m$. This
function is implemented in a distributed environment, thus ensuring a response
before new data is available.  After $L$ and $p$ parameters are found, VECM is
built and used to forecast the next data point.

\subsection{Model comparison} \label{sec:random}
We compared our proposal, in terms of performance, with the naive forecast of
the random walk model and ARIMA. It is still difficult to outperform the random
walk model for standard econometric forecasting models  despite its
simplicity, see \cite{lo2011}. The random walk model is defined as:

\begin{equation}
\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
\label{rwmodel}
\end{equation}

The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for
the random walk model is defined as:
\begin{equation}
\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
\end{equation}
\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.

On the other hand, ARIMA is widely used to forecast returns in finance, see
\cite{tsay2005}. A process can be modelled as an ARIMA$(p,d,q)$ model if
$\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time
series $\mathbf{y}_t$,  we get an ARMA$(p,q)$. Since we are modelling returns,
we used $d=1$.

\subsection{Evaluation methods} \label{sec:evaluation}

Forecast performance was evaluated using two different methods which are
frequently used in finance:
\begin{description}
\item
{\bf MSE},  Mean Square Error measures the distance between forecasts
and the true values and large deviations from the true value have a
large impact due to squaring forecast error.
\begin{equation}\label{eq:MSE}
\text{MSE} = 
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
\end{equation}
\item {\bf $U$-statistic}, the Theil's $U$-statistic, presented by
\cite{theil1966}, is a unit free measure obtained as the ratio between the root
MSE (RMSE) of a model and the RMSE of the naive random walk model. A
$U$-statistic less than 1 implies the performance is better than the naive
model.
\end{description}

\section{Experimental results}
\label{sec:results}
\subsection{Data} \label{sec:unitroot}
AVECM tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from
\cite{Dukascopy2014}, a free database which gives access to the Swiss Foreign
Exchange marketplace.

The tests were done using 10-second frequency from ask prices from the 11th to
the 15th of August 2014. Since one day corresponds to 8640 data points and we
used 5 days of data, we have 43,200 data points in total.

\subsection{Unit root tests} \label{sec:unitroottests}

Before running the tests, we firstly checked whether the time series were I(1)
using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
\cite{mackinnon2010} presented critical values for rejection of hypothesis of a
unit root: -2.62 (1\%), 1.94 (5\%) and 1.62 (10\%).
\begin{table*}[ht]
\label{tab:adf}
\centering
\begin{tabular}{llllll}
\toprule
{Variable} & {ADF(1)} & {ADF(2)} & {ADF(3)} & {ADF(4)} & {ADF(5)}\\ 
\midrule
EURUSD &  -0.052   & -0.054  & -0.054  & -0.054  & -0.054  \\
GBPUSD &  -0.744  & -0.784  & -0.805  & -0.837  & -0.846  \\
USDCHF &  -0.476   & -0.493  & -0.493  & -0.495  & -0.502  \\
USDJPY &  0.357   & 0.360  & 0.360  & 0.367  & 0.367  \\
$\Delta$ EURUSD & -128.4*  & -128.4*  & -96.85* & -89.12*   & -89.12*\\
$\Delta$ GBPUSD & -131.4*  & -112.7*  & -102.5* & -92.86*   & -88.29*\\
$\Delta$ USDCHF & -127.8*  & -127.8*  & -96.94* & -88.82*   & -80.79*\\
$\Delta$ USDJPY & -135.1*  & -135.1*  & -101.2* & -101.2*   & -101.2*\\
\bottomrule
\addlinespace[1ex]
\multicolumn{6}{l}{ \textsuperscript{*} Indicates significance at 1\% level} \\
\multicolumn{6}{l}{ \textsuperscript{**} Indicates significance at 5\% level} \\
\multicolumn{6}{l}{ \textsuperscript{***} Indicates significance at 10\% level} \\
\multicolumn{6}{l}{MacKinnon critical values for rejection of hypothesis of a unit root are:}\\
\multicolumn{6}{l}{ -2.62 (1\%), -1.94 (5\%) and -1.62 (10\%)}\\
\multicolumn{6}{l}{ADF($d$) Augmented Dickey-Fuller test with lag $d$} 
\end{tabular}
\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-second
frequency.}
\end{table*}
Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
test in their level form considering different lags, but they rejected it with
their first differences. This means that all of them are I(1) time series and we
are allowed to use VECM and therefore our proposed AVECM.

\subsection{Performance accuracy} \label{sec:performance}

Algorithms AVECM, ARIMA and the naive random walk were tested using four days of
data (from the 12th to the 15th of August 2014).  For AVECM we considered
different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10,
50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and
$p$. Possible values for $L$ were in the interval $[2,14]$ hours and $p$ can
have values in between $[1,5]$. Best AVECM performance was compared against
ARIMA and the random walk model.  Table~\ref{tab:stats} shows the out-of-sample
performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of
both measures we found that AVECM is superior to ARIMA and the naive random walk
model. We also included the p-value that proves that the difference in the MSE
is significant at the 99\% significance level in three of the four currency
rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and
ARIMA are superior to the naive random walk model and that our proposal is also
superior to ARIMA.

\begin{table}[!htbp]
\caption{AVECM performance}
\label{tab:stats}
\begin{center}
\begin{tabular}{ccccccc}

& \multicolumn{3}{l}{MSE} & &
\multicolumn{2}{c}{$U$-statistic} \\ 
\hhline{~---~--} \\
 &
AVECM & ARIMA & p-value & &
AVECM & ARIMA \\

\hline
 EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.2509 e-12 & & 0.6863 & 0.7108\\
 GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.9519 e-02 & & 0.6866 & 0.7025\\
 USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.8999 e-14 & & 0.6803 & 0.7091\\
 USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.8536 e-05 & & 0.6964 & 0.7057\\
 \end{tabular}
 \end{center}
\end{table}

\subsection{Parallel implementation} \label{sec:parallel}

To determine $L$ and $p$ based on maximising the percentage of cointegration
requires use of the Johansen method which is a computationally expensive
routine. This procedure is done by the function
get\_best\_params($Ls,ps,m,\mathbf{Y}$) shown in algorithm \ref{alg:AVECM}.  In
order to improve the execution time of this search, our proposal included a
parallel search of VECM parameters using high performance computing.  The main
objective was to obtain a response before a new data arrived in the following 10
seconds.

The Johansen method is already programmed in the Python Statsmodels
library, see \cite{seabold2010}, and the parallel implementation was developed using
MPI in Python.  We chose MPI because it allows large-scale parallel applications
with wide portability to be built, being able to run in large clusters or on
local computers.  We tested our proposal in a cluster with 2 servers Xeon
E5-2667 (2.90GHz) of 24 cores each (48 cores in total) and 24GB RAM.  In order
to compare serial and parallel execution times in AVECM, we set parameter
$it=100$ in algorithm \ref{alg:AVECM}.

The $L$ parameter was always chosen between 2 and 14 hours and $p$ always took
values between 1 and 5. Parameter $nparams$ represents the number of pairs
($L$,$p$) used to maximise the percentage of cointegration.

Execution time depends directly on $L$, $p$ and $nparams$ used, since they
determine the size of matrix $\mathbf{A}$ (see equation \ref{eq:Amatrix}) and
therefore affect the OLS function execution time.  Therefore, if we try more
combinations of $L$ and $p$ (increasing $nparams$) the serial algorithm will
take longer. For this financial time series we are interested in execution times
below 10 seconds  (the time series frequency).

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\textwidth]{img/Fig3}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}

The superior figure in \ref{fig:extimes} shows that best performance accuracy
measurements are achieved in times near or below 10 seconds in the parallel
version. Contrarily, serial times are higher, above 10 seconds in most cases.
Figure~\ref{fig:extimes} also shows the speed-up in computing time for AVECM
which is near 9X if we use more than 25 parameters ($nparams$).

Execution times do not consider the loading data time, just that of finding best
parameters and matrix operations. The time MPI spends transferring data and
synchronising processes is about two seconds independently of the number of
processes considered. Execution times were measured using the Time python
library.
